### 2019-09-27 paper/paper.tex
Add files via upload

Appealing to the Central Limit Theorem and the Principle of Maximum Entropy provides strong justification for assuming that the distribution of the sample mean, $\bar{x}_j$, for each sample is Gaussian, 
\begin{equation}
\bar{x}_j \sim N(\mu_j,\sigma_j^2/n_j ), \; \;    j=1,2,
\label{eq:xbar}
\end{equation}
where $\mu_j$ and $\sigma_j^2$ are the unknown population mean and variance of $x_j$, and $n_j$ is the number of observations in sample $x_j$. Adopting uninformative priors for the mean and variance leads to conditional posterior densities,
\begin{equation}
\mu_j | \sigma_j,x_j \sim N(\bar{x}_j, s_j^2),
\label{eq:muj}
\end{equation}
\begin{equation}
\sigma_j^2|\mu_j,x_j \sim IG((n_j-1)/2,(n_js_j^2)/2).
\label{eq:sigmaj}
\end{equation}
where $s_j^2$ is the sample variance. This is as far as we need to go analytically.  The rest of the problem can be solved numerically (though one more step is easily taken in this case, leading to a Student-t marginal posterior density for $\mu_j$ with posterior mean $\bar{x}_j$, variance $s_j^2/n$, and degrees of freedom $n_j-1$).
\vskip 6pt
Obtaining $M$ draws from these conditional distributions provides pseudo-samples from the marginal posteriors for $\mu$ and $\sigma^2$ (or drawing directly from the Student-t marginal posterior for $\mu$).  We then obtain an MCMC pseudo-sample of size $M$ from the posterior distribution of $\delta$, $\delta^{(m)} = \mu_1^{(m)} - \mu_2^{(m)},$ $m = 1,...,M$.  In this way, posteriors for any function of the parameters are available, such as differences in differences and ratios. For example, we often wish to compare two treatments to placebo, then to each other, which is accomplished by obtaining $\Delta^{(m)} = \delta_1^{(m)} - \delta_2^{(m)}$. While the analytical distribution of $\Delta$ is unknown and asymptotic approximations require unrealistic assumptions,  MCMC sampling allows the exact small sample posterior density to be approximated arbitrarily closely as $M$ is increased. The rest of the analysis requires only knowledge of basic statistics: plotting the density of the MCMC draws, computing summary statistics, and testing hypotheses.   Note that this addresses the Behrens-Fisher problem of testing the difference between the means of two populations when the variances of the two populations are not assumed either known or equal\cite{Ramos2010}.  Differing sample sizes for each group are also allowed.  Correlation across samples and other model extensions are straightforward.
\vskip 6pt
To test the hypotheses in (\ref{eq:hoh1}) we evaluate the posterior density ratio, PDR, (or `posterior odds') against $H_0$, by evaluating the MCMC posterior density for $\delta$ at the value in the null hypothesis and at the mode.  For treatment samples $x_1$ and $x_2$ the PDR is,
\begin{equation}
PDR(\delta = 0|x_1, x_2) = \frac{p(\delta=\delta_{MAP}|x_1,x_2)}{p(\delta=0|x_1,x_2)},
\label{eq:odds}
\end{equation}
where $\delta_{MAP}$ is the maximum a posteriori estimate of $\delta$.
We can also compute posterior tail probabilities (one-sided `Bayesian $p$-values'), from the MCMC sample,
\begin{equation}
p\text{-value} = \min{\left[P(\delta \le 0|x_1, x_2),P(\delta \ge 0|x_1, x_2)\right]},
\label{eq:pval}
\end{equation} 
$$P(\delta \le 0|x_1, x_2) =  \frac{\sum_{m=1}^M I(\delta^{(m)} \le 0) }  {M},$$
$$P(\delta \ge 0|x_1, x_2) =  \frac{\sum_{m=1}^M I(\delta^{(m)} \ge 0) }  {M},$$
where the indicator function $I(.) = 1$ if the condition is true, $0$ otherwise. These same formulas can be used to evaluate hypotheses concerning other quantities of interest, such as $\Delta$. The PDR for joint hypotheses (suppose $\delta$ is a vector of parameters) can readily be evaluated using Rao-Blackwellization\cite{Mills2018,Mills2019}.

@@

A Bayesian hierarchical model (BHM) provides a flexible setting for modeling data that allows for heterogeneity across individuals or groups. If homogeneity across studies is assumed, the data can be combined via Bayesian updating as in the sequential analysis in the previous section.  For categorical data, all that is needed are the number of occurrences, $s_i$, in $n_i$ subjects in each study, $i$, of $N$ studies. The posterior distribution is then 
%\begin{equation}
%\begin{split}
%p(\theta | s_1,...,s_N, n_1,...,n_N,a,b) = \\ \textnormal{Beta}(\sum_{i=1}^N s_i %+a,\sum_{i=1}^N(n_i - s_i) + b),
%\end{split}
%\label{eq:pbeta}
%\end{equation}
\begin{equation}
%\begin{split}
p(\theta | s, n,a,b) = \\ \textnormal{Beta}( s + a, n - s + b),
%\end{split}
\label{eq:pbeta}
\end{equation}
where $s = \sum_{i=1}^N s_i$ and $n = \sum_{i=1}^N n_i$, with $a=b=1$ for a uniform prior for $\theta$.
\vskip 6pt
Alternatively, BHMs can allow for heterogeneity across studies due to different trial sites, different treatments, such as different medications, etc., and the same Bayesian machinery can be employed to compare a variety of treatments relative to placebo. For the categorical data examined herein there are two common BHM specifications: the Beta-Gamma\cite{Kruschke2014}, and the Logistic-Normal\cite{Mcelreath2015}.  In the following, we adopted the Beta-Gamma BHM specification as it directly provides intuitively understandable parameter estimates, whereas the logistic specification is more difficult to interpret (though is better suited when covariates are included).
\vskip 6pt
The complete Beta-Gamma BHM modeling framework is illustrated in Figure \ref{fig:bhm} for the case of comparing two sets of RCTs for different treatments (sertraline vs. fluoxetine) with placebo groups in each RCT. Each RCT in Figure \ref{fig:bhm} has a Binomial likelihood for $s_i$ successes in $n_i$ trials with probability of success $\theta_i$ (row 3).  A common Beta prior with mode $\omega$ and precision parameter $K$ is assigned to each $\theta_i$ (row 2), and a Beta and Gamma hierarchical prior are assigned to $\omega$ and $K$ respectively (row 1). The hyperparameters $a,b,d$ and $r$ are selected to represent a priori knowledge about the distributions of $\omega$ and $K$ (relatively uninformative priors are usually assigned unless other information is available).
\vskip 6pt
The package Turing.jl provides Hamiltonian Monte Carlo (HMC) and MCMC algorithms to obtain posterior density MCMC chains from an large variety of models. The Julia macros provide a flexible and intuitive modeling and inference framework. The model specification in Turing.jl closely matches how one would write the model mathematically, e.g. for the Beta-Gamma BHM (Turing code examples for the Logistic-Normal are available at github.com/TuringLang/Turing.jl and github.com/StatisticalRethinkingJulia):
\begin{lstlisting}[language = Julia]
# Turing.jl BHM for binomial trials
using Turing, MCMCChains
@model binomial_trials(s,n) = begin
	g = length(n)  # number of groups
	# hierarchical priors
	ω ~ Beta(2,3)
	K ~ Gamma(10,1/0.05)
	a = ω*K + 1.0
	b = (1.0 - ω)*K + 1.0
	# priors for each occurrence rate
	θ = Array{Real}(undef, g)
	for k in 1:g
		θ[k] ~ Beta(a,b)
	end
	# likelihood
	for i in 1:g
		s[i] ~ Binomial(n[i],θ[i])
	end
end
\end{lstlisting}

Only a few more lines of code are needed to obtain posterior samples from the No U-Turn (NUTS) or HMC samplers in Turing.jl, and then evaluate the hypothesis of no mean difference between two groups:
\begin{lstlisting}[language = Julia]
using BayesTesting

# treatment
ctrt = mapreduce(c->sample(binomial_trials(s2, n2),
	NUTS(5000,1000,0.65)),chainscat,  1:5)
	# for the HMC sampler use HMC(5000, 0.05, 10)

# placebo
cpbo = mapreduce(c->sample(binomial_trials(s1, n1),
	NUTS(5000,1000,0.65)), chainscat,  1:3)

@@

# compute mean, SD, 0.95 prob. interval,
# PDR odds and tail prob.
[mean(δ) std(δ)]
quantile(δ,[0.025,0.5,0.975])
[mcodds(δ, h0=0.0) bayespval(δ)]
\end{lstlisting}

>>>>

Appealing to the Central Limit Theorem and the Principle of Maximum Entropy provides strong justification for assuming that the distribution of the sample mean, $\bar{x}_j$, for each sample is Gaussian, 
\begin{equation}
\bar{x}_j \sim N(\mu_j,\sigma_j^2/n_j ), \; \;    j=1,2,
\label{eq:xbar}
\end{equation}
where $\mu_j$ and $\sigma_j^2$ are the unknown population mean and variance of $x_j$, and $n_j$ is the number of observations in sample $x_j$. Adopting uninformative priors for the mean and variance leads to conditional posterior densities,
\begin{equation}
\mu_j | \sigma_j,x_j \sim N(\bar{x}_j, \sigma_j^2 /n),
\label{eq:muj}
\end{equation}
\begin{equation}
\sigma_j^2|x_j \sim IG((n_j-1)/2,(n_js_j^2)/2) = Inv-\chi^2(n_j-1,s_j^2).
\label{eq:sigmaj}
\end{equation}
where $s_j^2$ is the sample variance \cite{Gelman2004}. This is as far as we need to go analytically.  The rest of the problem can be solved numerically (though one more step is easily taken in this case, leading to a Student-t marginal posterior density for $\mu_j$ with posterior mean $\bar{x}_j$, variance $s_j^2/n$, and degrees of freedom $n_j-1$).
\vskip 6pt
Obtaining $M$ draws from these conditional distributions provides pseudo-samples from the marginal posteriors for $\mu$ and $\sigma^2$ (or drawing directly from the Student-t marginal posterior for $\mu$).  We then obtain an MCMC pseudo-sample of size $M$ from the posterior distribution of $\delta$, $\delta^{(m)} = \mu_1^{(m)} - \mu_2^{(m)},$ $m = 1,...,M$.  In this way, posteriors for any function of the parameters are available, such as differences in differences and ratios. For example, we often wish to compare two treatments to placebo, then to each other, which is accomplished by obtaining $\Delta^{(m)} = \delta_1^{(m)} - \delta_2^{(m)}$. While the analytical distribution of $\Delta$ is unknown and asymptotic approximations require unrealistic assumptions,  MCMC sampling allows the exact small sample posterior density to be approximated arbitrarily closely as $M$ is increased. The rest of the analysis requires only knowledge of basic statistics: plotting the density of the MCMC draws, computing summary statistics, and testing hypotheses.   Note that this addresses the Behrens-Fisher problem of testing the difference between the means of two populations when the variances of the two populations are not assumed either known or equal\cite{Ramos2010}.  Differing sample sizes for each group are also allowed.  Correlation across samples and other model extensions are straightforward.
\vskip 6pt
To test the hypotheses in (\ref{eq:hoh1}) we evaluate the posterior density ratio, PDR, (or `posterior odds') against $H_0$, by evaluating the MCMC posterior density for $\delta$ at the value in the null hypothesis and at the mode.  For treatment samples $x_1$ and $x_2$ the PDR is,
\begin{equation}
PDR(\delta = 0|x_1, x_2) = \frac{p(\delta=\delta_{MAP}|x_1,x_2)}{p(\delta=0|x_1,x_2)},
\label{eq:odds}
\end{equation}
where $\delta_{MAP}$ is the maximum a posteriori estimate of $\delta$.
We can also compute posterior tail probabilities (one-sided `Bayesian $p$-values'), from the MCMC sample,
\begin{equation}
p\text{-value} = \min{\left[P(\delta \le 0|x_1, x_2),P(\delta \ge 0|x_1, x_2)\right]},
\label{eq:pval}
\end{equation} 
$$P(\delta \le 0|x_1, x_2) =  \frac{\sum_{m=1}^M I(\delta^{(m)} \le 0) }  {M},$$
$$P(\delta \ge 0|x_1, x_2) =  \frac{\sum_{m=1}^M I(\delta^{(m)} \ge 0) }  {M},$$
where the indicator function $I(.) = 1$ if the condition is true, $0$ otherwise. These same formulas can be used to evaluate hypotheses concerning other quantities of interest, such as $\Delta$. The PDR for joint hypotheses (suppose $\delta$ is a vector of parameters) can readily be evaluated using Rao-Blackwellization\cite{Mills2018,Mills2019}.

@@

A Bayesian hierarchical model (BHM) provides a flexible setting for modeling data that allows for heterogeneity across individuals or groups. If homogeneity across studies is assumed, the data can be combined via Bayesian updating as in the sequential analysis in the previous section.  For categorical data, all that is needed are the number of occurrences, $s_i$, in $n_i$ subjects in each study, $i$, of $N$ studies. The posterior distribution is then 
%\begin{equation}
%\begin{split}
%p(\theta | s_1,...,s_N, n_1,...,n_N,a,b) = \\ \textnormal{Beta}(\sum_{i=1}^N s_i %+a,\sum_{i=1}^N(n_i - s_i) + b),
%\end{split}
%\label{eq:pbeta}
%\end{equation}
\begin{equation}
%\begin{split}
p(\theta | s, n,a,b) = \\ \textnormal{Beta}( s + a, n - s + b),
%\end{split}
\label{eq:pbeta}
\end{equation}
where $s = \sum_{i=1}^N s_i$ and $n = \sum_{i=1}^N n_i$, with $a=b=1$ for a uniform prior for $\theta$.
\vskip 6pt
Alternatively, BHMs can allow for heterogeneity across studies due to different trial sites, different treatments, such as different medications, etc., and the same Bayesian machinery can be employed to compare a variety of treatments relative to placebo. For the categorical data examined herein there are two common BHM specifications: the Beta-Gamma\cite{Kruschke2014}, and the Logistic-Normal\cite{Mcelreath2015}.  In the following, we adopted the Beta-Gamma BHM specification as it directly provides intuitively understandable parameter estimates, whereas the logistic specification is more difficult to interpret (though is better suited when covariates are included).
\vskip 6pt
The complete Beta-Gamma BHM modeling framework is illustrated in Figure \ref{fig:bhm} for the case of comparing two sets of RCTs for different treatments (sertraline vs. fluoxetine) with placebo groups in each RCT. Each RCT in Figure \ref{fig:bhm} has a Binomial likelihood for $s_i$ successes in $n_i$ trials with probability of success $\theta_i$ (row 3).  A common Beta prior with mode $\omega$ and precision parameter $K$ is assigned to each $\theta_i$ (row 2), and a Beta and Gamma hierarchical prior are assigned to $\omega$ and $K$ respectively (row 1). The hyperparameters $a,b,d$ and $r$ are selected to represent a priori knowledge about the distributions of $\omega$ and $K$ (relatively uninformative priors are usually assigned unless other information is available).
\vskip 6pt
The package Turing.jl provides Hamiltonian Monte Carlo (HMC) and MCMC algorithms to obtain posterior density MCMC chains from an large variety of models. The Julia macros provide a flexible and intuitive modeling and inference framework. The model specification in Turing.jl closely matches how one would write the model mathematically, e.g. for the Beta-Gamma BHM (Turing code examples for the Logistic-Normal are available at github.com/TuringLang/Turing.jl and github.com/StatisticalRethinkingJulia):
\begin{lstlisting}[language = Julia]
using Turing, MCMCChains
@model binomial_trials(s,n) = begin
	g = length(n)  # number of groups
	# hierarchical priors
	ω ~ Beta(2,3)
	K ~ Gamma(10,1/0.05)
	a = ω*K + 1.0
	b = (1.0 - ω)*K + 1.0
	# priors for each occurrence rate
	θ = Array{Real}(undef, g)
	for k in 1:g
		θ[k] ~ Beta(a,b)
	end
	# likelihood
	for i in 1:g
		s[i] ~ Binomial(n[i],θ[i])
	end
end
\end{lstlisting}

A few more lines of code provide posterior samples from the No U-Turn (NUTS) sampler in Turing.jl, allowing evaluation of the no mean difference between two groups hypothesis:
\begin{lstlisting}[language = Julia]
using BayesTesting

# treatment
ctrt = mapreduce(c->sample(binomial_trials(s2, n2),
	NUTS(5000,1000,0.65)),chainscat,  1:5)
# placebo
cpbo = mapreduce(c->sample(binomial_trials(s1, n1),
	NUTS(5000,1000,0.65)), chainscat,  1:3)

@@

# compute summary stats, PDR odds, tail prob.
[mean(δ) std(δ)]
quantile(δ,[0.025,0.5,0.975])
[mcodds(δ, h0=0.0) bayespval(δ)]
\end{lstlisting}
